---
title: Lobster GS model calibration
format: html
editor: visual
---

## Calibration work-flow for Lobster GS model application

**Step 0.** Create reverse mode input files for each experimental dataset

**Step 1.** Run **app.R** to explore parameters to tune model to observations in reverse mode (Initial model exploration)

**Step 2.** Determine **initial** fV by fitting regression of Lipid vs Protein intakes across feeds for the experimental observation and run as inputs in forward mode

**Step 3.** Using initial fV and model inputs estimate the best parameter set for each feed (lowest error between pred and obs) for the selected parameter(s) to be tuned by using optimization or grid-search#

**Step 4.** Repeat Steps 0-3 for each experiment and make plot of pred vs obs across all model and data.

**Step 5.** Using best parameter sets for all experiments, run model in forward mode to predict N waste vs % Protein in Feed (and could predict other outputs such as growth, respiration etc., to explore further)

## **We are working with 3 datasets**

1.  Wirtz et al. 2022
2.  Landman et al 2021a
3.  Landman et al. 2021b

## The code below is for Wirtz et al 2022

```{r}
rm(list=ls()) 
library(conflicted)
library(shiny)
library(plotly) 
library(DT)
library(tidyverse)
library(pbapply)
library(ggrepel) 
library(purrr)
```

### Step 0 - get data

ensure tau and eta values correspond to this tau 0.04; eta 0.00001 in the GS_modelfunctions.R file

```{r}
#EXPERIMENT 1 -> wirtzinput.csv and wirtzobs.csv
source("GS_model_functions.R")
fvinput<-read.csv("input_Data/wirtzinput.csv")
fvinput$Protein_intake<-NA
fvinput$Lipid_intake<-NA
input<-read.csv("input_Data/wirztobs.csv")  
```

### Step 1 - run app

```{r}
source("app.R") #shiny_gs() ## interactively, this indicates kstarN as candidate parameter
```

### Step 2 - get fV values

use linear regression or exact observed intake values for each feed

```{r}
coefs<-lm(data=input,Lipid_intake ~Protein_intake-1) 
predLI<-predict(coefs) #Predicts corresponding lipid intake for each value of protein intake
```

#### predict fV for each datapoint

since we have observed values of intakes,we use the code below to gather fV in GS forward mode

*Note: we are assuming model defaults of kstarN and phi before estimating candidate parameters below*

```{r}
out<-pbapply(X=as.matrix(input[,c(2,3,5,6)]),1,GSforwardmodel,kstarN=0.9,phi=0) 
fV<-c(out[[1]]$fV,out[[2]]$fV,out[[3]]$fV,out[[4]]$fV)
# add to obsinput
input$fV<-fV
```

*IMPORTANT: check fV values for all feeds here. If any of them are 0, please change the optimisation function to optim (c(0.9,0) instead of the model default optim (c(0.9,0.5) we assume from Anderson et al 2020*

### Step 3 - get best parameter set for each feed type using optimisation function

```{r}
# Control parameters for optimization
control_params <- list(maxit = 50000) 

FMest<-optim(c(0.9,0.5),getError,input=input,feed="FM",paramname="both",method="L-BFGS-B",upper=c(1.0,1.0),lower=c(0.00001,0.00001), control = control_params)
# fV= 0 here so I changed the value of phi for KM
KMest<-optim(c(0.9,0.0),getError,input=input,feed="KM",paramname="both",method="L-BFGS-B",upper=c(1.0,1.0),lower=c(0.00001,0.00001),control = control_params) 

SBMest<-optim(c(0.9,0.5),getError,input=input,feed="SBM",paramname="both",method="L-BFGS-B",upper=c(1.0,1.0),lower=c(0.00001,0.00001),control = control_params) 

# fV= 0 here so I changed the value of phi for SWM
SWMest<-optim(c(0.9,0),getError,input=input,feed="SWM",paramname="both",method ="L-BFGS-B",upper=c(1.0,1.0),lower=c(0.00001,0.00001),control = control_params) 
pars<-rbind(FMest$par, KMest$par,SBMest$par,SWMest$par)

#check to see if these make sense 
feed=c("FM","KM","SBM","SWM")
input$pred_Protein_intake<-NA
input$pred_Lipid_intake<-NA
for (i in 1:4) { 
finput<-data.frame(G=input[input$Feed==feed[i],"G"],fV=input[input$Feed==feed[i],"fV"],
 betaV=input[input$Feed==feed[i],"betaV"],betaH=input[input$Feed==feed[i],"betaH"], 
 input[input$Feed==feed[i],"Lipid_intake"],input[input$Feed==feed[i],"Protein_intake"])
 input[i,c(8:9)]<-GSreversemodel(finput,kstarN = pars[i,1],phi = pars[i,2]) } 
#adding kstarN and phi to input dataframe 
input$kstarN<-pars[,1]
input$phi<-pars[,2]

#saving input here to use as dataframe for goodness of fit plot (fig 5 of MS)
saveRDS(input,"fig5wirtzdata.RDS") #we will call on this later to plot figure 5. the outputs here are different to the "newwirtzinput.RDS" file which is used as inputs for GS forward model 
```

#### Create full array of inputs for fV 0 –\>1 to predict intake array (rails)

```{r}
## replace matrix 10 times for each value of fV 0:1
M <- as.matrix(input[,-1]) 
input_array <- matrix(rep(t(M), 1000), ncol=dim(M)[2], byrow=TRUE) 
colnames(input_array) <- names(input[,-1])
input_array <- data.frame(Feed=input[,1], input_array) 

# replace fVs with values from 0.001 to 1 in steps of 0.001
input_array[,"fV"] <- rep(seq(0.001, 1, 0.001), each=nrow(M))

# write over observed Protein and Lipid intake values with predicted ones
for (i in 1:dim(input_array)[1]) {
    input_array[i, c("pred_Protein_intake", "pred_Lipid_intake")] <- unlist(GSreversemodel(
        input_array[i, ],
        kstarN = input_array[i, "kstarN"],
        phi = input_array[i, "phi"]
    ))
}

```

#### get calibrated predictions for plotting figure 3

```{r}
input$pred_Protein_intake<-NA
input$pred_Lipid_intake<-NA 
for (i in 1:4){ input[i,c(8:9)]<-GSreversemodel(as.matrix(input[i,-1]),kstarN = input[i,"kstarN"],phi = input[i,"phi"]) }
```

### Figure 4: Wirtz plot intake arrays

plotting input arrays, comparing calibrated vs uncalibrated vs expt

```{r}
library(viridisLite)
library(viridis) 
library(ggplot2) 
library(ggrepel)

wirtzplot <- ggplot(input, aes(x = Protein_intake, y = Lipid_intake, group = G)) +
  geom_point(size = 3.5, aes(color = G)) +
  geom_point(size = 6, shape = "X", data = input, aes(x = pred_Protein_intake, y = pred_Lipid_intake, color = G)) +
  geom_abline(slope = unlist(coefs[1]), intercept = 0, linetype = 2) +
  geom_path(data = input_array, aes(x = pred_Protein_intake, y = pred_Lipid_intake, color = G)) +
  scale_color_viridis_c() + 
  lims(x = c(0, 0.11), y = c(0, 0.06)) +
  labs(size = 4, 
       x = expression("Protein intake mol C mol C"^"-1" * " day"^"-1"),
       y = expression("Lipid intake mol C mol C"^"-1" * " day"^"-1"),
       color = "growth rate") +
  theme(axis.title = element_text(size = 14)) +
  theme(legend.position = c(1, 1), 
        legend.justification = c("right", "top"), 
        legend.box.just = "right", 
        legend.margin = margin(6, 6, 6, 6)) +
  geom_text_repel(size = 3, data = input, aes(label = Feed)) +
  annotate("text", x = 0.008, y = 0.008, label = "fV=0.37", angle = 0) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"))

wirtzplot <- wirtzplot + theme(aspect.ratio = 0.75)
#note that since SWM phi value is also predicted as 0, WE DONT NEED THE DOTTED LINE 
print(wirtzplot)

#ggsave("manuscript_plots/wirtzplot.png",wirtzplot)
```

### we also need to predict waste output here before moving on to the next experiment

```{r}
names(input)[8]<-"Protein_intake" 
names(input)[9]<-"Lipid_intake" 
saveRDS(input,"newwirtzinput.RDS")
```

### now we do wirtz waste plot in forward mode for Figure 3

```{r}
wirtzfwdmode<-readRDS("newwirtzinput.RDS")
rownames<-wirtzfwdmode$Feed #select which columns are needed to run in forward mode 
wirtzfwdmode<-as.matrix(wirtzfwdmode[,c(5,6,8,9,10,11)]) 
rownames(wirtzfwdmode)<-rownames # below the columns being selected needed to be fixed: 
out2<-pbapply(X=wirtzfwdmode,1,GSforwardmodel) # test run holding kstarN and phi constant
test2<-pbapply(X=wirtzfwdmode[,c(1,2,3,4)],1,GSforwardmodel,kstarN=0.9,phi=0)
out2<-pbapply(X=wirtzfwdmode,1,GSforwardmodel) 

out3<-c(out2[[1]]$Pellet_N,out2[[2]]$Pellet_N,out2[[3]]$Pellet_N,out2[[4]]$Pellet_N)
```

#### TEST: compare predictions with test

```{r}
test3<-c(test2[[1]]$Pellet_N,test2[[2]]$Pellet_N,test2[[3]]$Pellet_N,test2[[4]]$Pellet_N) # waste looks the same

outg<-c(out2[[1]]$Growth_G,out2[[2]]$Growth_G,out2[[3]]$Growth_G,out2[[4]]$Growth_G) 
testg<-c(test2[[1]]$Growth_N,test2[[2]]$Growth_N,test2[[3]]$Growth_N,test2[[4]]$Growth_N) # growth is very different
#printoutput 
print(out2)
#now convert to df 
wirtzfwdmode<-data.frame(wirtzfwdmode) 
wirtzfwdmode$Nwaste<-out3
wirtzfwdmode$CP_feed<-c(63.75,65.1,60.05,60.55)
wirtzfwdmode$Feed<-rownames
#saveRDS(wirtzfwdmode, "wirtzfwdmodedata.RDS") #dont need these but saving in case we need to look at the dataset
```

#### Plotting

remember to save parameter values and outputs above (for other metabolic outputs) from above

### FIGURE 3: panel plot (3 rows).

##### A: C:N ratio of feeds vs growth rate day-1; plot showing relationship between feed C:N ratio and growth

##### B: C:N ratio of feeds vs N waste output mol N mol C-1 day-1; Relationship between C:N ratio and faecal N waste output

##### C: Differences in C:N ratio of feeds and consumer vs differences in N waste

```{r}
# Import dataset - some data manipulation is needed after obtaining model outputs
panelplot <- read.csv("input_data/wirtzpanelplotv1.csv", encoding = "UTF-8")

# Load necessary libraries
library(ggplot2)
library(grid)
library(wesanderson)
library(ggpubr)
library(ggrepel)
library(ggtext)

# Plot 1
plot1 <- ggplot(panelplot, aes(x = C_N_ratio_feed, y = growthG_model., color = feed)) +
  geom_point(size = 6) +
  geom_point(shape = "x", size = 9, data = panelplot, aes(x = C_N_ratio_feed, y = growthG_expt)) +
  geom_text_repel(aes(label = feed), size = 6, box.padding = 0.5, point.padding = 0.5, max.overlaps = Inf, min.segment.length = 0.1) +
  theme(legend.position = "none",
        axis.title = element_text(size = 16),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.text = element_text(size = 14),
        axis.line = element_line(colour = "black"))+
  labs(size = 11, x = expression("C:N ratio of feeds"), y = expression(atop("growth rate", paste("day"^"-1")))) +
  scale_color_manual(values = wes_palette("GrandBudapest1"))

# Plot 2
plot2 <- ggplot(panelplot, aes(x = C_N_ratio_feed, y = n_waste_model, color = feed)) +
  geom_point(size = 6) +
  geom_point(shape = "x", size = 9, data = panelplot, aes(x = C_N_ratio_feed, y = n_waste_conv)) +
  geom_text_repel(aes(label = feed), size = 6, box.padding = 0.75, point.padding = 0.75, max.overlaps = Inf, min.segment.length = 0.1) +
  theme(legend.position = "none",
        axis.title = element_text(size = 16),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.text = element_text(size = 14),
        axis.line = element_line(colour = "black")) +
  labs(size = 11, x = expression("C:N ratio of feeds"), y = expression(atop("N waste output", paste("mol N mol C"^"-1"*"day"^"-1")))) +
  scale_color_manual(values = wes_palette("GrandBudapest1"))

# Plot 3
plot3 <- ggplot(panelplot, aes(x = C_N_ratio_diff_con_feed, y = n_waste_percentage, color = feed)) +
  geom_point(size = 6) +  
  geom_text_repel(aes(label = feed), size = 6, box.padding = 0.75, point.padding = 0.75, max.overlaps = Inf, min.segment.length = 0.1)  + 
  theme(legend.position = "none",
        axis.title = element_text(size = 16),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.text = element_text(size = 14),
        axis.line = element_line(colour = "black")) +  coord_cartesian(xlim =c(0,4.5),ylim = c(0, 200)) + 
  labs(size = 11, x = expression("Differences in C:N ratio of feeds and consumer"), y = expression(atop("Percentage differences in N waste", paste("(model prediction- conventional approach)")))) +
  scale_color_manual(values = wes_palette("GrandBudapest1"))

# Arrange plots
wirtzwasteplot <- ggarrange(plot1, plot2, plot3, nrow = 3)

# Save the plot as pdf
# Open the PDF device
cairo_pdf("wirtzwasteplot.pdf", width = 10, height = 16.5)

# Print the combined plot to the device
print(wirtzwasteplot)

# Close the PDF device
dev.off()
```

#### EXPERIMENT 2 -\> Go back to model functions file, change tau and eta values,

#### Use the code below for landman2021a calibration

#### for Landman 2021a : tau 0.04; eta 0.00002

```{r}
source("GS_model_functions.R") 
fvinput<-read.csv("input_data/landman2021input.csv") #your input file with G,fV range from 0-1, betaV and betaH if available 
fvinput$Protein_intake<-NA
fvinput$Lipid_intake<-NA # "wirztobs.csv", "landman2021aobs.csv", "landman2021bobs.csv" 
input<-read.csv("input_data/landman2021aobs.csv")    
```

## Step 1: Run app

```{r}
 source("app.R") #shiny_gs() ## interactively, this indicates kstarN as candidate parameter
```

## Step 2 - get fV values

```{r}
coefs<-lm(data=input,Lipid_intake ~Protein_intake-1) 
predLI<-predict(coefs) #PREDICT LIPID INTAKE coefs predLI
```

use linear regression or exact observed intake values for each feed

#determining fV for each datapoint when kstarN=0.9,phi=0

```{r}
out<-pbapply(X=as.matrix(input[,c(2,3,5,6)]),1,GSforwardmodel,kstarN=0.9,phi=0) 
fV<-c(out[[1]]$fV,out[[2]]$fV,out[[3]]$fV,out[[4]]$fV,out[[5]]$fV,out[[6]]$fV)
#add to obsinput
input$fV<-fV
```

*IMPORTANT: check fV values for all feeds here. If any of them are 0, please change the optimisation function to optim (c(0.9,0) instead of the model default optim (c(0.9,0.5) we assume from Anderson et al 2020*

## Step 3 - get best parameter set for each feed type

```{r}
#for more than one variable to estimate need to use optim
D1est<-optim(c(0.9,0.5),getError,input=input,feed="D1",paramname="both",method="L-BFGS-B",upper=c(1.0,1.0),lower=c(0.00001,0.00001)) 
D2est<-optim(c(0.9,0.5),getError,input=input,feed="D2",paramname="both",method="L-BFGS-B",upper=c(1.0,1.0),lower=c(0.00001,0.00001))
D3est<-optim(c(0.9,0.5),getError,input=input,feed="D3",paramname="both",method="L-BFGS-B",upper=c(1.0,1.0),lower=c(0.00001,0.00001)) 
D4est<-optim(c(0.9,0.5),getError,input=input,feed="D4",paramname="both",method ="L-BFGS-B",upper=c(1.0,1.0),lower=c(0.00001,0.00001)) 

#fV for D5 was predicted as 0, so have changed the optim below to reflect that 
D5est<-optim(c(0.9,0.5),getError,input=input,feed="D5",paramname="both",method ="L-BFGS-B",upper=c(1.0,1.0),lower=c(0.00001,0.00001)) 
D6est<-optim(c(0.9,0),getError,input=input,feed="D6",paramname="both",method ="L-BFGS-B",upper=c(1.0,1.0),lower=c(0.00001,0.00001)) 
pars<-rbind(D1est$par,D2est$par,D3est$par,D4est$par,D5est$par,D6est$par)
```

### check to see if these make sense

```{r}
feed=c("D1","D2","D3","D4","D5","D6")
input$pred_Protein_intake<-NA
input$pred_Lipid_intake<-NA 
for (i in 1:6){ finput<-data.frame(G=input[input$Feed==feed[i],"G"],fV=input[input$Feed==feed[i],"fV"],betaV=input[input$Feed==feed[i],"betaV"],betaH=input[input$Feed==feed[i],"betaH"],input[input$Feed==feed[i],"Lipid_intake"],input[input$Feed==feed[i],"Protein_intake"])
input[i,c(8:9)]<-GSreversemodel(finput,kstarN = pars[i,1],phi = pars[i,2]) }

#add pars kstarN and phi to dataframe
input$kstarN<-pars[,1]
input$phi<-pars[,2]

#replace matrix 10 times for each value of fV 0:1
M<-as.matrix(input[,-1]) 
input_array<-matrix(rep(t(M),10),ncol=dim(M)[2],byrow=TRUE) 
colnames(input_array)<-names(input[,-1]) 
input_array<-data.frame(Feed=input[,1],input_array) # replace fVs with 0.1 to 1 instead of calibrated values 
#create full array of inputs with varying fV form 0:1; #check to see if these make sense
input_array[,"fV"]<-rep(seq(0.1,1,0.1),each=6)

#write over observed Protein and Lipid intake values with predicted ones
for (i in 1:dim(input_array)[1]) input_array[i,c("pred_Protein_intake","pred_Lipid_intake")]<-unlist(GSreversemodel(input_array[i,],kstarN=input_array[i,"kstarN"],phi=input_array[i,"phi"]))

```

#### get predicted calibrated values for plotting

```{r}
# Initialize columns
input$pred_Protein_intake <- NA
input$pred_Lipid_intake <- NA 

# Loop to fill the predictions
for (i in 1:6) { 
  input[i, c(8:9)] <- GSreversemodel(as.matrix(input[i, -1]), kstarN = input[i, "kstarN"], phi = input[i, "phi"])
}

#saving input as dataframe here for goodness of fit plot 
saveRDS(input,"fig5landman1data.RDS")


# Plotting observed vs predicted intakes with intake rails
library(ggplot2)
library(ggrepel)
library(viridis)

landman2021aplot <- ggplot(input, aes(x = Protein_intake, y = Lipid_intake, group = G)) + 
  scale_color_continuous(type = "viridis") + 
  geom_point(size = 3.5, aes(color = G)) + 
  geom_point(size = 5, shape = "X", data = input, aes(x = pred_Protein_intake, y = pred_Lipid_intake, color = G)) + 
  geom_abline(slope = unlist(coefs[1]), intercept = 0, linetype = 2) + 
  geom_path(data = input_array, aes(x = pred_Protein_intake, y = pred_Lipid_intake, color = G)) + 
  lims(x = c(0, 0.11), y = c(0, 0.06)) + 
  labs(size = 4, 
       x = expression("Protein intake mol C mol C"^"-1"~day^"-1"), 
       y = expression("Lipid intake mol C mol C"^"-1"~day^"-1"), 
       color = "growth rate") + 
  theme(axis.title = element_text(size = 14)) + 
  theme(legend.position = c(1,1), legend.justification = c("right", "top"), legend.box.just = "right", legend.margin = margin(6, 6, 6, 6)) + 
  geom_text_repel(size = 3, data = input, aes(label = Feed)) + 
  annotate("text", x = 0.025, y = 0.007, label = "fV=0.18", angle = 0) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black"))

landman2021aplot<- landman2021aplot + theme(aspect.ratio = 0.75)
print(landman2021aplot)


#ggsave("manuscript_plots/landman2021aplot.png",landman2021aplot)

# Saving input data
# Renaming predicted intake columns
names(input)[8] <- "Protein_intake" 
names(input)[9] <- "Lipid_intake" 

# Save dataframe as RDS
saveRDS(input, "newlandman1input.RDS")
```

#### Supplementary figure 1: Landman et al 2021a wasteplot

```{r}
library(wesanderson) 
landman1fwdmode<-readRDS("newlandman1input.RDS") 
rownames<-landman1fwdmode$Feed
landman1fwdmode<-as.matrix(landman1fwdmode[,c(5,6,8,9,10,11)])
rownames(landman1fwdmode)<-rownames
out4<-pbapply(X=landman1fwdmode,1,GSforwardmodel)
#save output here if needed for other metabolic outputs
saveRDS(out4, "landman1fwdmodeoutputs.RDS")
out5<-c(out4[[1]]$Pellet_N,out4[[2]]$Pellet_N,out4[[3]]$Pellet_N,out4[[4]]$Pellet_N,out4[[5]]$Pellet_N,out4[[6]]$Pellet_N)
landman1fwdmode$Nwaste<-out5 
landman1fwdmode$CP_feed<-c(60.5,59.77,60.57,61.34,61.02, 61.82) 
landman1fwdmode<-data.frame(landman1fwdmode) 
l1waste<-ggplot(landman1fwdmode, aes(x = CP_feed, y = Nwaste, color = rownames)) +
  geom_point(size = 4) +
  geom_text_repel(size = 4, color = "black", data = landman1fwdmode, aes(label = rownames)) +
  scale_color_manual(values = wes_palette("BottleRocket1")) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "black"),
    legend.position = "none",
    axis.title = element_blank() # This removes the axis labels
  ) +
  lims(x = c(52, 67), y = c(0, 0.010))

#+ labs(size=8,x=expression("C:N ratio in feed"),y=expression("Nitrogenous waste output mol N mol C"^"-1"*"day"^"-1"))

#saving plot as a grob 
g6 <- ggplotGrob(l1waste)
```

#### EXPERIMENT 3 -\> landman2021binput.csv

#### go back to model functions file, change tau and eta values tau 0.02; eta 0.00001

#### Step 0: Import datasets

```{r}
source("GS_model_functions.R")
 #get input data - ""landman2021binputs.csv" 
fvinput<-read.csv("input_data/landman2021binputs.csv") #your input file with G,fV range from 0-1, betaV and betaH if available 
fvinput$Protein_intake<-NA
fvinput$Lipid_intake<-NA # "wirztobs.csv", "landman2021aobs.csv", "landman2021bobs.csv" 
input<-read.csv("input_data/landman2021bobs.csv")
```

#### Step 1: Run app

```{r}
source("app.R") #shiny_gs() ## interactively, this indicates kstarN as candidate parameter
```

#### Step 2 - get fV values

use linear regression or exact observed intake values for each feed

```{r}
coefs<-lm(data=input,Lipid_intake ~Protein_intake-1)
predLI<-predict(coefs) #PREDICT LIPID INTAKE coefs predLI
```

#### or for each data point

```{r}
out<-pbapply(X=as.matrix(input[,c(2,3,5,6)]),1,GSforwardmodel,kstarN=0.9,phi=0) 
fV<-c(out[[1]]$fV,out[[2]]$fV,out[[3]]$fV,out[[4]]$fV,out[[5]]$fV,out[[6]]$fV,out[[7]]$fV)
# add to obsinput
input$fV<-fV
```

*IMPORTANT: check fV values for all feeds here. If any of them are 0, please change the optimisation function to optim (c(0.9,0) instead of the model default optim (c(0.9,0.5) we assume from Anderson et al 2020*

#### Step 3 - get best parameter set for each feed type \# for more than one variable to estimate need to use optim

```{r}
BM0est<-optim(c(0.9,0.5,0.6,0.6),getError,input=input,feed="BM0%",paramname="withbetas",method="L-BFGS-B",upper=c(1.0,1.0,1.0,1.0),lower=c(0.00001,1e-10,1e-10,1e-10)) 
BM16est<-optim(c(0.9,0.5,0.6,0.6),getError,input=input,feed="BM1.6%",paramname="withbetas",method="L-BFGS-B",upper=c(1.0,1.0,1.0,1.0),lower=c(0.00001,1e-10,1e-10,1e-10)) 
BM31est<-optim(c(0.9,0.5,0.6,0.6),getError,input=input,feed="BM3.1%",paramname="withbetas",method="L-BFGS-B",upper=c(1.0,1.0,1.0,1.0),lower=c(0.00001,1e-10,1e-10,1e-10)) 
BM63est<-optim(c(0.9,0.5,0.6,0.6),getError,input=input,feed="BM6.3%",paramname="withbetas",method ="L-BFGS-B",upper=c(1.0,1.0,1.0,1.0),lower=c(0.00001,1e-10,1e-10,1e-10)) 
BM125est<-optim(c(0.9,0.5,0.6,0.6),getError,input=input,feed="BM12.5%",paramname="withbetas",method ="L-BFGS-B",upper=c(1.0,1.0,1.0,1.0),lower=c(0.00001,1e-10,1e-10,1e-10)) 
BM250est<-optim(c(0.9,0.5,0.6,0.6),getError,input=input,feed="BM25%",paramname="withbetas",method ="L-BFGS-B",upper=c(1.0,1.0,1.0,1.0),lower=c(0.00001,0.00001,0.00001,0.00001)) 
BMHSest<-optim(c(0.9,0.5,0.6,0.6),getError,input=input,feed="BMHS",paramname="withbetas",method ="L-BFGS-B",upper=c(1.0,1.0,1.0,1.0),lower=c(0.00001,0.00001,0.00001,0.00001)) 
pars<-rbind(BM0est$par,BM16est$par,BM31est$par,BM63est$par,BM125est$par,BM250est$par,BMHSest$par)

#adding columns for parameters
colnames(pars)<-c("kstarN","phi","betaV","betaH") 
rownames(pars)<-input$Feed
#replace input with calibrated values 
input[,"betaV"]<-pars[,"betaV"] 
input[,"betaH"]<-pars[,"betaH"] 
input$phi<-input$kstarN<-NA
input[,"kstarN"]<-pars[,"kstarN"]
input[,"phi"]<-pars[,"phi"]

#replace matrix 10 times for each value of fV 0:1

M<-as.matrix(input[,-1]) 
input_array<-matrix(rep(t(M),10),ncol=dim(M)[2],byrow=TRUE) 
colnames(input_array)<-names(input[,-1]) 
input_array<-data.frame(Feed=input[,1],input_array) 
# replace fVs with 0.1 to 1 instead of calibrated values 
input_array[,"fV"]<-rep(seq(0.1,1,0.1),each=7)

#check to see if these make sense

#create full array of inputs with varying fV form 0:1

#write over observed Protein and Lipid intake values with predicted ones
for (i in 1:dim(input_array)[1])input_array[i,c("pred_Protein_intake","pred_Lipid_intake")]<-unlist(GSreversemodel(input_array[i,],kstarN=input_array[i,"kstarN"],phi=input_array[i,"phi"],betaV=input_array[i,"betaV"],betaH=input_array[i,"betaH"]))

#get predicted calibrated values for plotting

#feed=c("BM0","BM16","BM31","BM63","BM125","BM250","BMHS") 
input$pred_Protein_intake<-NA
input$pred_Lipid_intake<-NA
for (i in 1:7){ input[i,c(10:11)]<-GSreversemodel(as.matrix(input[i,-1]),kstarN = pars[i,"kstarN"],phi = pars[i,"phi"],betaV = pars[i,"betaV"],betaH = pars[i,"betaH"]) }


#saving input as dataframe for goodness of fit plot 
saveRDS(input,"fig5landman2data.RDS")


```

Figure 3: Intake arrays/ rails

```{r}
#plotting
# Load necessary libraries
library(ggplot2) 
library(ggrepel)
library(readr)
library(viridis)

# Ensure input is a data frame
input <- as.data.frame(input)

# Plotting observed vs predicted intakes with intake rails
landman2021bplot <- ggplot(input, aes(x = Protein_intake, y = Lipid_intake, color = G)) + 
  scale_color_continuous(type = "viridis") + 
  geom_point(size = 3.5) + 
  geom_point(data = input, size = 6, shape = "X", aes(x = pred_Protein_intake, y= pred_Lipid_intake, z = G)) + 
  geom_abline(slope = unlist(coefs[1]), intercept = 0, linetype = 2) + geom_line(data = input_array, aes(x = pred_Protein_intake, y = pred_Lipid_intake, group = G)) + 
  lims(x = c(0, 0.11), y = c(0, 0.06)) + 
  labs(size = 4, 
       x = expression("Protein intake mol C mol C"^"-1"~day^"-1"), 
       y = expression("Lipid intake mol C mol C"^"-1"~day^"-1"), 
       color = "growth rate") + 
  theme(axis.title = element_text(size = 14)) + 
  theme(legend.position = c(1,1), legend.justification = c("right", "top"), legend.box.just = "right", legend.margin = margin(6, 6, 6, 6)) + 
  geom_text_repel(size = 3, data = input, aes(label = Feed), color = "black") + 
  annotate("text", x = 0.01, y = 0.006, label = "fV=0.25", angle = 0) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black"))

landman2021bplot<-landman2021bplot+ theme(aspect.ratio = 0.75)
#Print the plot
print(landman2021bplot)
ggsave("manuscript_plots/landman2021bplot.png", landman2021bplot)
```

renaming pred_protein_intake and pred_Lipid_intake as IV and IH for saving df to run in forward mode

```{r}
names(input)[10]<-"Protein_intake" 
names(input)[11]<-"Lipid_intake" #save df as RDS just in case you need to go back and ref and you've cleared the environment 
library(readr)
saveRDS(input,"newlandman2input.RDS") #wasteplot 
landman2fwdmode<-readRDS("newlandman2input.RDS") 
rownames<-landman2fwdmode$Feed
landman2fwdmode<-as.matrix(landman2fwdmode[,c(5,6,8,9,10,11)])
rownames(landman2fwdmode)<-rownames
out6<-pbapply(X=landman2fwdmode,1,GSforwardmodel)
out7<-c(out6[[1]]$Pellet_N,out6[[2]]$Pellet_N,out6[[3]]$Pellet_N,out6[[4]]$Pellet_N,out6[[5]]$Pellet_N,out6[[6]]$Pellet_N,out6[[7]]$Pellet_N) 
landman2fwdmode$Nwaste<-out7
landman2fwdmode$CP_feed<-c(62.31,62.68,61.19,60.82,61.44,61.49,54.31) 
landman2fwdmode<-as.data.frame(landman2fwdmode)

##plotting C:N ratio in feed vs Nitrogenous waste output
library(ggplot2) 
library(ggrepel)
library(wesanderson)

#Your original ggplot code without theme and limits
l2waste <- ggplot(landman2fwdmode, aes(x = CP_feed, y = Nwaste, color = rownames)) + 
  geom_point(size = 4) + 
  scale_color_manual(values = wes_palette("BottleRocket1")) + 
  geom_text_repel(size = 4, color = "black", aes(label = rownames)) + 
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.title = element_blank(), # This removes the axis labels
    legend.position = "none"
  ) + 
  lims(x = c(52, 67), y = c(0, 0.015))

#+ labs(x = expression("C:N ratio in feed"), y = expression("Nitrogenous waste output mol N mol C"^"-1"*"day"^"-1"))
print(l2waste) 
#save l2waste as a grob 
g7 <- ggplotGrob(l2waste)
```

# Supplementary information figure 1: N waste predictions for study 2 and 3

**Combining landman 1 and landman 2 in a single plot window**

```{r}
# Load necessary libraries
library(ggpubr)
library(ggplot2)
library(grid)

# Assuming g6 and g7 are already defined ggplot objects
# Combine the plots
g8 <- ggarrange(g6, g7, nrow = 1, ncol = 2)

# Annotate the combined plot
annotated_g8 <- annotate_figure(
  g8,
  left = textGrob(expression("N waste output mol C"~mol~C^{-1}~day^{-1}), rot = 90, vjust = 1, gp = gpar(cex = 1.3)),
  bottom = textGrob(expression("Crude protein percentage(%) in feed"), gp = gpar(cex = 1.3))
)

# Save the annotated plot to TIFF format
tiff("supfig1.tiff", units="in", width=16.5, height=6.3, res=300)
print(annotated_g8)  # Print the annotated plot to the TIFF device
dev.off()  # Close the TIFF device
```

#### Figure 4: Protein and lipid intake arrays #Panel plot (3 columns).

#### A. Wirtz et al; B. Landman et al 2021a; C. Landman et al 2021b

```{r}
library(gridExtra)
library(grid)
library(png)
library(grid)

library(cowplot)
library(grid)
library(cowplot)
library(grid)

# Define your plots
g1 <- as_grob(wirtzplot)
g2 <- as_grob(landman2021aplot)
g3 <- as_grob(landman2021bplot)

# Combine plots with labels
combined_plot <- plot_grid(
  g1, g2, g3,
  ncol = 3,
  labels = c("A", "B", "C"),
  label_size = 16,  # Size of the labels
  label_x = c(0.05, 0.35, 0.65),  # x positions for labels
  label_y = 1.05  # y position for labels
)

# Open the PDF device
cairo_pdf("combined_plot.pdf", width = 16.5, height = 6.3)

# Print the combined plot to the device
print(combined_plot)

# Close the PDF device
dev.off()

```

#### FIGURE 5: Goodness of fit tests

#### predicted vs observed protein and lipid values

```{r}
#combine the dataframes from all 3 experiments
df1<-readRDS("fig5wirtzdata.RDS")
df2<-readRDS("fig5landman1data.RDS")
df3<-readRDS("fig5landman2data.RDS")
df_combined <-bind_rows(df1,df2,df3)
saveRDS(df_combined, "goodnessoffitdata.RDS")
fitdata<-readRDS("goodnessoffitdata.RDS")
```

#### Plotting

```{r}
#predicted vs observed we want two panels:
#a) all values across all experiments predicted for Lipid intake vs all values across all experiments observed Lipid intakes, with a 1:1 line on the plot
library(ggplot2)
library(ggrepel)
#fitdata<-read.csv("fitdata.csv")
#readRDS("goodnessoffit.RDS")
proteinplot<-ggplot(fitdata,aes(x=pred_Protein_intake,y=Protein_intake))+geom_point()+ geom_abline(intercept=0,linetype=2)+geom_text_repel(size=5,data=fitdata,aes(label = Feed))+
  theme(axis.title = element_text(size = 14),panel.grid.major = element_blank(),panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black"))+ 
  labs(size=8,x=expression("Predicted crude protein intake mol C mol C"^"-1"*"day"^"-1"),y=expression("Observed crude protein intake mol C mol C"^"-1"*"day"^"-1"),par(cex.axis = 1.3))

#Calculating rsquared
x<-fitdata$pred_Protein_intake
y<-fitdata$Protein_intake
proteinpred<-lm(y~x)
summary(proteinpred)$r.squared


#b)all values across all experiments predicted for Protein intake vs all values across all experiments observed Protein intakes, with a 1:1 line on the plot
lipidplot<-ggplot(fitdata,aes(x=pred_Lipid_intake,y=Lipid_intake))+geom_point()+ geom_abline(intercept=0,linetype=2)+geom_text_repel(size=5,data=fitdata,aes(label = Feed))+ 
theme(axis.title = element_text(size = 14),panel.grid.major = element_blank(),panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black"))+ 
  labs(size=8,x=expression("Predicted lipid intake mol C mol C"^"-1"*"day"^"-1"),y=expression("Observed lipid intake mol C mol C"^"-1"*"day"^"-1"),par(cex.axis = 1.3))

#calculating rsquared
a<-fitdata$pred_Lipid_intake
b<-fitdata$Lipid_intake
lipidpred<-lm(b~a)
summary(lipidpred)$r.squared

#combine above plots in single window 
library(ggpubr)
library(ggplot2)
library(grid)
predvsobserved<-ggarrange(proteinplot,lipidplot,ncol = 2,nrow=1,labels = c("A","B"))

#save the above plot as a tiff file. 
#tiff("fig5.tiff", units = "in", width = 16.5, height = 6.3, res = 300)
print(predvsobserved)  # Print again before saving
#dev.off()
```

#### Identifying outliers in lipid prediction- v2 of plot excluding SWM

```{r}
# Load necessary libraries
library(ggplot2)
library(ggrepel)
library(ggpubr)

# Function to identify outliers based on IQR
identify_outliers <- function(data, column) {
  Q1 <- quantile(data[[column]], 0.25, na.rm = TRUE)
  Q3 <- quantile(data[[column]], 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR
  data[[column]] < lower_bound | data[[column]] > upper_bound
}

# Identify outliers for Protein intake
protein_outliers <- identify_outliers(df_combined, "Protein_intake")
protein_outliers_df <- df_combined[protein_outliers, ]
print("Protein Intake Outliers:")
print(protein_outliers_df)

# Identify outliers for Lipid intake
lipid_outliers <- identify_outliers(df_combined, "Lipid_intake")
lipid_outliers_df <- df_combined[lipid_outliers, ]
print("Lipid Intake Outliers:")
print(lipid_outliers_df)

# Combine outliers from both Protein and Lipid intakes
outliers_combined <- Reduce(`|`, list(protein_outliers, lipid_outliers))

# Remove outliers, keeping 'BMHS'
df_clean <- df_combined[!(outliers_combined & df_combined$Feed != 'none'), ]

# Optional: Check how many rows were excluded
print(paste("Number of rows excluded:", nrow(df_combined) - nrow(df_clean)))

# Protein intake plot
proteinplot <- ggplot(df_clean, aes(x = pred_Protein_intake, y = Protein_intake)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = 2) +
  geom_text_repel(size = 5, aes(label = Feed)) +
  theme(axis.title = element_text(size = 14),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black")) +
  labs(x = expression("Predicted crude protein intake mol C mol C"^"-1"*"day"^"-1"),
       y = expression("Observed crude protein intake mol C mol C"^"-1"*"day"^"-1"),
       par(cex.axis = 1.3))

# Calculating R-squared for protein intake
protein_r_squared <- summary(lm(Protein_intake ~ pred_Protein_intake, data = df_clean))$r.squared
print(paste("Protein intake R-squared:", protein_r_squared))

# Lipid intake plot
lipidplot <- ggplot(df_clean, aes(x = pred_Lipid_intake, y = Lipid_intake)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = 2) +
  geom_text_repel(size = 5, aes(label = Feed)) +
  theme(axis.title = element_text(size = 14),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black")) +
  labs(x = expression("Predicted lipid intake mol C mol C"^"-1"*"day"^"-1"),
       y = expression("Observed lipid intake mol C mol C"^"-1"*"day"^"-1"),
       par(cex.axis = 1.3))

# Calculating R-squared for lipid intake
lipid_r_squared <- summary(lm(Lipid_intake ~ pred_Lipid_intake, data = df_clean))$r.squared
print(paste("Lipid intake R-squared:", lipid_r_squared))

# Combine plots in a single window
predvsobserved <- ggarrange(proteinplot, lipidplot, ncol = 2, nrow = 1, labels = c("A", "B"))
print(predvsobserved)

# Save the plot
#tiff("fig5_outliersremoved.tiff", units = "in", width = 16.5, height = 6.3, res = 300)
print(predvsobserved)  # Print again before saving
#dev.off()
```
